\section{What is coalgebraic about algebraic effects and handlers?}

Besides strong connection with algebras, algebraic effects also have some relatively unexplored coalgebraic features. The main objects of our study will be \emph{comodels} of an algebraic theory. Let us start with an abstract categorical definition and work back from there to a useful characterization. Those not interested in categorical manipulations may skip to Section~\ref{sec:comodels} and take that as a starting point.

\subsection{Dualizing models of algebraic theories}
\label{sec:dualizing-models}

Recall that for a given category $\category{C}$, the dual category $\opcat{\category{C}}$ has the same objects as $\category{C}$, whereas its morphisms go in the other direction: morphisms between $A$ and $B$ in $\opcat{\category{C}}$ correspond exactly to morphisms $B \to A$ in $\category{C}$, which is why we will write them as $A \from B$, keeping objects in the same place and reversing the arrows. We use bold letters to write $\text{\textbf{op}}$, the standard notation for a dual category, in order to distinguish it from $\op{}$, the standard notation for an algebraic operation.

Take $\category{C}$ to be any category with finite \emph{coproducts}. We roughly follow Plotkin \& Power~\cite{power} and define the category of $\theory{T}$-comodels $\ComodC{C}{T}$ and $\theory{T}$-cohomomorphisms to be $\opcat{(\ModC{\opcat{C}}{T})}$, i.e. the dual of $\theory{T}$-models in the dual of $\category{C}$. Let us express these in more familiar terms.

First, consider the objects of $\ComodC{C}{T}$. Since a category has the same objects as its dual, these correspond to objects of $\ModC{\opcat{C}}{T}$, which are given by:
%
\begin{enumerate}
\item An object $|W|$ in $\opcat{\category{C}}$, called the \emph{carrier}. Again, a category has the same objects as its dual, these correspond to objects $|W|$ in $\category{C}$.
\item For each operation symbol $\op{i}$ a morphism in $\opcat{\category{C}}$
  %
  \begin{equation*}
    \sem{\op{i}}_W : \underbrace{|W| \times \cdots \times |W|}_{\arity{i}} \from |W|,
  \end{equation*}
  which validates all the equations of the theory (we will return to this in a moment). However, products in $\opcat{\category{C}}$ correspond to coproducts in $\category{C}$ (which is why we required $\category{C}$ to have finite coproducts), thus the above family of morphisms in $\opcat{\category{C}}$ corresponds to the family of morphisms
  %
    \begin{equation*}
    \sem{\op{i}}^W : |W| \to \underbrace{|W| + \cdots + |W|}_{\arity{i}},
  \end{equation*}
  in $\category{C}$, which we call \emph{cooperations}. We abbreviate an $n$-ary coproduct $|W| + \cdots + |W|$ by $n \cdot |W|$.
\end{enumerate}
Since the category $\ComodC{C}{T} = \opcat{(\ModC{\opcat{C}}{T})}$ has the same objects as its dual $\ModC{\opcat{C}}{T}$, such objects are exactly $\theory{T}$-comodels.

A family of cooperations on a model validates the equations of the theory~$\theory{T}$ if for every equation
%
\begin{equation*}
  x_1, \ldots, x_k \mid \ell = r
\end{equation*}
%
in~$\equations{T}$, the maps $\sem{x_1, \ldots, x_k \mid \ell}^W$ and $\sem{x_1, \ldots, x_k \mid r}^W$ are equal. Interpretations of $\Sigma$-terms are, of course, defined dually. A $\Sigma$-term in context
%
\begin{equation*}
  x_0, \ldots, x_{k-1} \mid t
\end{equation*}
%
is interpreted by a map
%
\begin{equation*}
  \sem{x_0, \ldots, x_{k-1} \mid t}^I : |I| \to k \cdot |I|,
\end{equation*}
%
as follows:
%
\begin{enumerate}
\item the variable $x_i$ is interpreted as the $i$-th injection,
  %
  \begin{equation*}
    \sem{x_0, \ldots, x_{k-1} \mid  x_i}^I = \iota_i : |I| \to k \cdot |I|,
  \end{equation*}
\item a compound term in context
  %
  \begin{equation*}
    x_0, \ldots, x_{k-1} \mid \op{i}(t_1, \ldots, t_{\arity{i}})
  \end{equation*}
  %
  is interpreted as the composition of maps
  %
  \begin{equation*}
    \xymatrix@+6em{
      {|I|} \ar[r]^{\sem{\op{i}}^I}
      &
      {{\arity{i}} \cdot |I|} \ar[r]^{[\sem{t_0}_I, \ldots, \sem{t_{\arity{i}}}_I]}
      &
      {k \cdot |I|}
    }
  \end{equation*}
\end{enumerate}
%
where for morphisms $f_i \colon A_i \to B$, we define the morphism $[f_1, \dots, f_k] \colon A_1 + \cdots + A_k \to B$ by $[f_1, \dots, f_k](\iota_j(x)) = f_j(x)$.

Finally, morphisms between comodels $W$ and $V$ in $\ComodC{C}{T}$ are exactly $\theory{T}$-homomorphisms between $V$ and $W$ in its dual $\ModC{\opcat{C}}{T}$. These in turn correspond to morphisms $\phi \colon |V| \from |W|$ in $\opcat{C}$ that commute with operations: for every operation symbol $\op{i}$ of~$\theory{T}$, we have
%
\begin{equation*}
  \phi \circ \sem{\op{i}}_V = \sem{\op{i}}_W \circ \underbrace{(\phi, \ldots, \phi)}_{\arity{i}}
\end{equation*}
in the category $\opcat{C}$. If we state this in terms of $\category{C}$, we get a morphism $\phi \colon |W| \to |V|$ that commutes with cooperations: for every operation symbol $\op{i}$ of~$\theory{T}$, we have
%
\begin{equation*}
  \sem{\op{i}}_V \circ \phi = \underbrace{[\phi, \ldots, \phi]}_{\arity{i}} \circ \sem{\op{i}}_W.
\end{equation*}

Let us now move to the category of sets, and while doing this, let us also switch to operations with parameters and a general arity. We see that in sets, $k \cdot |W| = [k] \times |W|$, so an operation symbol $\op{}$ with parameter set~$P$ and arity~$A$ can be interpreted as a cooperation $P \cdot |W| \to A \cdot |W|$.

\subsection{Comodels of algebraic theories}
\label{sec:comodels}

A \emph{cointerpretation $I$ of a signature $\Sigma$} is given by:
%
\begin{enumerate}
\item a carrier set $|I|$,
\item for each operation symbol $\op{i}$ with parameter set~$P_i$ and arity~$A_i$,
  a map
  %
  \begin{equation*}
    \sem{\op{i}}^kI : P_i \times |I| \longrightarrow A_i \times |I|,
  \end{equation*}
  which we call a \emph{cooperation}.
\end{enumerate}
%
The interpretation $I$ may be extended to terms in contexts. A term
$x_1, \ldots, x_k \mid t$ is interpreted as a map
%
\begin{equation*}
  \sem{x_1, \ldots, x_k \mid t}_I : |I| \to k \cdot |I|
\end{equation*}
%
as follows (for those that skipped the first section, $k \cdot |I| = \underbrace{|I| + \cdots + |X|}k$):
%
\begin{enumerate}
\item the variable $x_i$ is interpreted as the $i$-th injection,
  %
  \begin{equation*}
    \sem{x_0, \ldots, x_{k-1} \mid  x_i}_I = \iota_i : |I| \to k \cdot |I|,
  \end{equation*}
\item a compound term in context
  %
  \begin{equation*}
    x_1, \ldots, x_k \mid \op{i}(p, \kappa)
  \end{equation*}
  %
  is interpreted as the map
  %
  \begin{align*}
    \sem{\op{i}(p, \kappa)}_I &: |I| \longrightarrow k \cdot |I| \\
    \sem{\op{i}(p, \kappa)}_I &:
      w \mapsto (a, \kappa)
      \sem{\op{i}}_I(p, \lam{a \in A_i} \sem{\kappa(a)}_I(\eta)),
  \end{align*}
\end{enumerate}

A \emph{$\Sigma$-equation} is a pair of $\Sigma$-terms $\ell$ and $r$ in context $x_1, \ldots, x_k$, written
%
\begin{equation*}
  x_1, \ldots, x_k \mid \ell = r.
\end{equation*}
%
Given an cointerpretation $I$ of signature $\Sigma$, we say that such an equation is \emph{valid} for~$I$ when the interpretations of $\ell$ and $r$ give the same map.

An \emph{algebraic theory $\theory{T} = (\signature{T}, \equations{T})$} is
given by a signature $\signature{T}$ and a collection of $\Sigma$-equations
$\equations{T}$. A \emph{$\theory{T}$-comodel} is a cointerpretation for
$\signature{T}$ which validates all the equations~$\equations{T}$.

\begin{example}
Take the signature~$\Sigma$ with a binary operation symbol $\mathsf{choose}$. A cointerpretation of $\Sigma$ is given by a set $S$ together with a map $\sem{\mathsf{choose}}^S \colon S \to 2 \times S$. This can be seen as a boolean stream, where on each call $\sem{\mathsf{choose}}^S$ produces a boolean and the remainder of the stream.
\end{example}

\begin{example}
A state mapping locations from $L$ into values from $V$ can be described by a signature~$\Sigma$ consisting of two operation symbols: $\mathsf{get}$ with parameter set $L$ and arity $V$, and a unary $\mathsf{set}$ with parameter set $L \times V$. The cointerpretation of $\Sigma$ amounts to a set $S$ with maps $\sem{\mathsf{get}} \colon L \times S \to V \times S$ and $\sem{\mathsf{set}} \colon (L \times V) \times S \to S$. One such set would be the set of functions $V^L$, where $\sem{\mathsf{get}}(\ell, s) = s(\ell)$ and $\sem{\mathsf{get}}((\ell, v), s) = s[\ell \mapsto v]$, i.e. a function that behaves as $s$, but maps $\ell$ to $v$.
\end{example}

\begin{example}
Output is represented with a unary operation symbol $\mathsf{print}$ with a parameter set $O$ representing the output alphabet. Cointerpretations of such signature are given by sets $W$ operation A state mapping locations from $L$ into values from $V$ can be described by a signature~$\Sigma$ consisting of two operation symbols: $\mathsf{get}$ with parameter set $L$ and arity $V$, and a unary $\mathsf{set}$ with parameter set $L \times V$. The cointerpretation of $\Sigma$ amounts to a set $S$ with maps $\sem{\mathsf{get}} \colon L \times S \to V \times S$ and $\sem{\mathsf{set}} \colon (L \times V) \times S \to S$. One such set would be the set of functions $V^L$, where $\sem{\mathsf{get}}(\ell, s) = s(\ell)$ and $\sem{\mathsf{get}}((\ell, v), s) = s[\ell \mapsto v]$, i.e. a function that behaves as $s$, but maps $\ell$ to $v$.
\end{example}



\begin{example}
Note that if $\Sigma$ contains a nullary operation symbol $\op : 0$, no nontrivial cointerpretation of $\Sigma$ exists, as that would entail a set $|W|$ and a map $|W| \to \emptyset$, which is possible only if $|W|$ is itself $\emptyset$.
\end{example}

\subsection{Using comodels for interpreting hardware}

We can think of comodels~$W$ as hardware on which operations are executed. The carrier~$|W|$ represents the state, while cooperations $|W| \times P \to |W| \times A$ take the initial state and a parameter, and produce the result and the altered state.

We can use this to execute programs, which, recall, are represented with models. Take a model~$M$ representing the software and a comodel~$W$ representing the hardware  on which it is run. Next, take an initial configuration $\conf{m}{w}$, where $m \in M$ is a program and $w \in W$ is the inital state. If $m$ is of the form $\return v$, our program has terminated with the resulting value $v$ in a final state~$w$. But if $m$ calls an operation and is of the form $\op{M}(p, \kappa)$, we can execute $\op{}$ on $W$. By applying the cooperation $\sem{\op}^W$ to the parameter $p$ and the world~$w$, we get a result~$a$ and a new world~$w'$. We pass $a$ to $\kappa$ to get a continuation, which we resume in the new state $w'$.

This relation induces an equivalence $\equiv$ on the set of configurations $M \times W$, given by:
\[
  \conf{\op{M}(p, \kappa)}{w} \equiv \conf{\kappa(a)}{w'} \quad\text{where $\op{}^W(p; w) = (a, w')$}
\]
We label the resulting quotient set $(M \times W) / \equiv$ by $M \otimes W$ and call it the \emph{tensor product}, because of its similarity to how scalars in a tensor product pass from one component to the other.

In case the theory~$\theory{T}$ is nontrivial, we must check that the relation is well-defined, for $\op{M}(p, \kappa)$ may be equivalent to some other $m' \in M$. Since $W$ is also a comodel, $\op{}^W$ obeys the same equations. We leave the proof as an exercise for the reader.

\subsection{Tensoring models and comodels of algebraic theories}
\label{sec:tensoring-models}

We have used comodels to represent the hardware that executes the operations that escape all handlers and reach the toplevel. Another frequent viewpoint is to represent it with a ``default handler'' which sits at the top and encompasses the whole executed computation. Let us compare the two representations and see why the former seems a better fit. First, we can show that any such hardware can indeed be simulated by a paramater-passing handler. For a comodel $W$ with cooperations, we can define a model $(X \times W)^W$ where the program computing values from $V$ can be ...
